---
title: "A1"
output:
  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
  html_notebook:
    toc: yes
    toc_depth: 2
bibliography: 'A1.bib'
---

# Downloading required packages

[@geometadb]
[@knitr]
[@edger]
[@biocmanager]
[@biomart]
```{r, message= FALSE}
if (!requireNamespace("GEOmetadb", quietly = TRUE))
  BiocManager::install("GEOmetadb")

if (!requireNamespace("knitr", quietly = TRUE))
  install.packages("knitr")

if (!require("edgeR", quietly = TRUE))
  BiocManager::install("edgeR")

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

if (!requireNamespace("biomaRt", quietly = TRUE))
  BiocManager::install("biomaRt")

```

# Calling required packages

```{r, message=FALSE}
library(BiocManager)
library(GEOmetadb)
library(knitr)
library(edgeR)
library(biomaRt)
```

# Download the data
## GSE152074 raw data supplemntary file downloaded

```{r, message=FALSE}
sfiles = getGEOSuppFiles('GSE152075')
fnames = rownames(sfiles)
# there is only one supplemental file
readData = read.table(fnames[1],header=TRUE, check.names = TRUE)
head(readData)

```

# Contact Information
```{r, message=FALSE}
gse <- getGEO("GSE152075",GSEMatrix=FALSE)

kable(data.frame(head(Meta(gse))), format = "html")
```

# Assess

Checking original data dimensions before cleaning
```{r, message=FALSE}
#How many genes do we have
dim(readData)


```

Add 1 to all values of data so later on when conducting log2(cpm) we can avoid negative infinity values. (Advised by Professor Isserlin)
```{r, message=FALSE}
readData <- readData + 1
```


Setting first column as gene id for future format purposes
```{r}
#Place rownames in first column for future format purposes
inter <- data.frame("HUGO" = rownames(readData))
geneData <- cbind(inter$HUGO, readData)
colnames(geneData)[1] <- "HUGO"
```



Check for any duplicates in gene ID
```{r, include=FALSE}
rawGeneCount <- sort(table(geneData$HUGO),decreasing = TRUE)

kable(rawGeneCount[
  which(rawGeneCount>1)], 
  format="html")
```

# Clean
Remove any outliers that does not have at least 1 read per million in n of the samples.
Denoting n as the smallest group of replicates which is the control group of 53.
Using n = 53 conduct the removal of low counts.
```{r}
#translate out counts into counts per millison using 
#the edgeR package function cpm
cpms = cpm(geneData[,2:485])
rownames(cpms) <- geneData[,1]
# get rid of low counts
keep = rowSums(cpms >1) >=53
geneData_exp_filtered = geneData[keep,]
nrow(geneData) - nrow(geneData_exp_filtered)
```

Remove version numbers if they exists on gene id(HUGO) column.
This makes it easier for mapping later on.
```{r}
geneData_exp_filtered[,1] <- gsub("\\.[0-9]", "", geneData_exp_filtered[,1])
```

# Map
```{r, message=FALSE}
#Mapping the name using biomatr
# list available gene annotation databases
bio <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# convert column of gene IDs to Hugo symbols
geneMapping <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                     mart = bio,
                     filters = "hgnc_symbol",
                     values = geneData_exp_filtered[,1])


```
Combine the mapped gene data to original data
```{r}
#Merge the data
mergedData <- merge(geneData_exp_filtered, geneMapping, by.x = 1, by.y = 2)
#remove duplicate rows in the gene data
mergedDataNoDup <- mergedData[!duplicated(mergedData[,1:485]),]



```

Check for duplicates
```{r}
summarized_gene_counts_filtered <- sort(table(mergedDataNoDup$HUGO),
                                        decreasing = TRUE)

kable(summarized_gene_counts_filtered[
  which(summarized_gene_counts_filtered>1)[1:10]], 
  format="html")
```


# Apply Normalization
Set seed for randomly sampling data for boxplot and density graphs before and after normalization.
```{r}
set.seed(12345)
randomSamplePOS <- sample(mergedDataNoDup[2:431], 10)
randomSampleNEG <- sample(mergedDataNoDup[432:485], 10)
randomSample <- cbind(randomSamplePOS,randomSampleNEG, mergedDataNoDup$ensembl_gene_id)

```

Define groups to use in normalization
```{r}
samples <- data.frame(lapply(colnames(randomSample[1:20]), 
        FUN=function(x){unlist(strsplit(x, 
                        split = "_"))[c(2,1)]}))
colnames(samples) <- colnames(randomSample[1:20])
rownames(samples) <- c("patients","cell_type")
samples <- data.frame(t(samples))

```



Applying TMM to data
```{r}

filtered_data_matrix <- as.matrix(randomSample[1:20])
rownames(filtered_data_matrix) <- randomSample$`mergedDataNoDup$ensembl_gene_id`
d = DGEList(counts=filtered_data_matrix, group=samples$cell_type)

d = calcNormFactors(d)

normalized_counts <- cpm(d)
```

## Not normalized plots
Boxplot for non-normalized data
```{r}
data2plot <- log2(cpm(randomSample[1:20]))
boxplot(data2plot, xlab = "Samples", ylab = "log2 CPM", 
        las = 2, cex = 0.5, cex.lab = 0.5,
        cex.axis = 0.5, main = "RNASeq Samples")
#draw the median on each box plot
abline(h = median(apply(data2plot, 2, median)), 
       col = "green", lwd = 0.6, lty = "dashed")
```
 
 Density plot for non-normalized data

```{r}
counts_density <- apply(log2(cpm(randomSample[1:20])), 
                        2, density)
  #calculate the limits across all the samples
    xlim <- 0; ylim <- 0
    for (i in 1:length(counts_density)) {
      xlim <- range(c(xlim, counts_density[[i]]$x)); 
      ylim <- range(c(ylim, counts_density[[i]]$y))
    }
    cols <- rainbow(length(counts_density))
    ltys <- rep(1, length(counts_density))
    #plot the first density plot to initialize the plot
    plot(counts_density[[1]], xlim=xlim, ylim=ylim, type="n", 
         ylab="Smoothing density of log2-CPM", 
         main="", cex.lab = 0.85)
    #plot each line
    for (i in 1:length(counts_density)) 
      lines(counts_density[[i]], col=cols[i], lty=ltys[i])
    #create legend
    legend("topright", colnames(data2plot), legend = colnames(randomSample[1:20]),
           col=cols, lty=ltys, cex=0.75, 
           border ="blue",  text.col = "green4", 
           merge = TRUE, bg = "gray90")
```

## Normalized Plots
Define groups for normalized data

Barplot for normalized data
```{r}
data2plotNorm <- log2(normalized_counts)
boxplot(data2plotNorm, xlab = "Samples", ylab = "log2 TMM", 
        las = 2, cex = 0.5, cex.lab = 0.5,
        cex.axis = 0.5, main = "RNASeq Samples")
#draw the median on each box plot
abline(h = median(apply(data2plotNorm, 2, median)), 
       col = "green", lwd = 0.6, lty = "dashed")
```

Density plot for normalized data
```{r}
countsDensityNorm <- apply(log2(normalized_counts), 
                        2, density)
  #calculate the limits across all the samples
    xlim <- 0; ylim <- 0
    for (i in 1:length(countsDensityNorm)) {
      xlim <- range(c(xlim, countsDensityNorm[[i]]$x)); 
      ylim <- range(c(ylim, countsDensityNorm[[i]]$y))
    }
    cols <- rainbow(length(countsDensityNorm))
    ltys <- rep(1, length(countsDensityNorm))
    #plot the first density plot to initialize the plot
    plot(countsDensityNorm[[1]], xlim=xlim, ylim=ylim, type="n", 
         ylab="Smoothing density of log2-TMM", 
         main="", cex.lab = 0.85)
    #plot each line
    for (i in 1:length(countsDensityNorm)) 
      lines(countsDensityNorm[[i]], col=cols[i], lty=ltys[i])
    #create legend
    legend("topright", colnames(data2plotNorm), legend = colnames(normalized_counts),  
           col=cols, lty=ltys, cex=0.75, 
           border ="blue",  text.col = "green4", 
           merge = TRUE, bg = "gray90")
```

Define groups for MDS plot
```{r}
samplesNorm <- data.frame(lapply(colnames(normalized_counts), 
        FUN=function(x){unlist(strsplit(x, 
                        split = "_"))[c(2,1)]}))
colnames(samplesNorm) <- colnames(normalized_counts)
rownames(samplesNorm) <- c("patients","cell_type")
samplesNorm <- data.frame(t(samplesNorm))

```


MDS for normalized data
```{r}
plotMDS(d, labels=rownames(samplesNorm),
  col = c("darkgreen","blue")[factor(samplesNorm$cell_type)])

```

# Final dataset
Formatting final dataset
```{r}
rownames(mergedDataNoDup) <- mergedDataNoDup$HUGO

dim(mergedDataNoDup)
```

**Important notes about code idea**
The normalization was done with only the 20 randomly sampled data because a boxplot, and density plot with over 400 samples was too chaotic.
The sample takes 10 control, and 10 case randomly from the whole dataset. 
Because of this only 20 samples were normalized for the analysis of plots.

**What are the control and test conditions of the dataset?**
The control is the NEG columns in the dataset that are not effected by SARS-COV-2, and the test is the POS are patients with SARS-COV-2 positive results from nasopharyngeal swabs.


**Why is the dataset of interest to you?**
My current BCB330Y1 research is about genetic association studies with SARS-Cov-2. I wanted to learn more about the topic, and use my time efficiently to expand my knowledge in this area.

**Were there expression values that were not unique for specific genes? How did you handle these?**
There were no cases of expression values that were not unique for specific genes. However there were cases where there were multiple Ensemblid for a single gene id given in the raw data.
Took care of this by only keeping one Ensemblid and discarding the rest.

**Were there expression values that could not be mapped to current HUGO symbols?**
There were about 1000 genes that did not map to HUGO symbols however out of 20,000 genes that passed the initial filtering process it is not significant.

**How many outliers were removed?**
14855 genes were removed with low counts that removed genes without at least 1 read per million in the sample. 

**How did you handle replicates?**
Replicates in the control and test conditions were all used in the data. However a majority were cut in the low counts filtering process.

**What is the final coverage of your dataset?**
The final dataset has 19291 genes, with 484 samples.53 control, and 431 case samples.



# References

All Code and statistics were heavily taken from **Lecture 4 - Exploring the data and Lecture 4 - Exploring the data and
basics of Normalization basics of Normalization** by Ruth Isserlin.


[@lieberman2020vivo]

